---
title: "FinalProject"
output: html_document
date: "2023-10-25"
---

## Required Packages

```{r}
library(tidyverse)
library(skimr)
library(MASS)
```

## Download Data

Data from Framingham Heart Study

```{r}
men_data = read.csv("fhs_men.csv")
women_data = read.csv("fhs_women.csv")

head(men_data)
head(women_data)

full_data = read.csv("framingham.csv")
```

## EDA

```{r}
library(ggplot2)

# age
ggplot(men_data, aes (x = age)) + geom_histogram(fill = "blue", alpha = 0.2) + ggtitle("Distribution of Men Ages") + xlab("Age")
ggplot(women_data, aes (x = age)) + geom_histogram(fill = "red", alpha = 0.2) + ggtitle("Distribution of Women Ages") + xlab("Age")

# education
ggplot(men_data, aes (x = as.factor(education), fill = as.factor(education))) + geom_bar() + ggtitle("Distribution of Men Education") + xlab("Education") + scale_fill_discrete(name = "Level of Education", labels = c("Higher Secondary", "Graduation", "Post-Graduation", "PHD"))
ggplot(women_data, aes (x = as.factor(education), fill = as.factor(education))) + geom_bar() + ggtitle("Distribution of Women Education") + xlab("Education") + scale_fill_discrete(name = "Level of Education", labels = c("Higher Secondary", "Graduation", "Post-Graduation", "PHD"))

#currentSmoker
ggplot(men_data, aes(x = "", fill = as.factor(currentSmoker))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Men Smokers") + theme_void() + scale_fill_discrete(name = "People", labels = c("Nonsmoker", "Smoker"))
ggplot(women_data, aes(x = "", fill = as.factor(currentSmoker))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Women Smokers") + theme_void() + scale_fill_discrete(name = "People", labels = c("Nonsmoker", "Smoker"))

#cigsPerDay
ggplot(men_data, aes (x = cigsPerDay)) + geom_histogram(fill = "green", alpha = 0.6, binwidth = 4) + ggtitle("Distribution of The Amount of Cigarettes Smoked by Male Smokers") + xlab("Cigarettes Per Day")
ggplot(women_data, aes (x = cigsPerDay)) + geom_histogram(fill = "pink", alpha = 0.6, binwidth = 4) + ggtitle("Distribution of The Amount of Cigarettes Smoked by Female Smokers") + xlab("Cigarettes Per Day")

#BPMeds
ggplot(men_data, aes(x = "", fill = as.factor(BPMeds))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Men on Blood Pressure Medication") + theme_void() + scale_fill_discrete(name = "People", labels = c("Not on BP Meds", "On BP Meds"))
ggplot(women_data, aes(x = "", fill = as.factor(BPMeds))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Women on Blood Pressure Medication") + theme_void() + scale_fill_discrete(name = "People", labels = c("Not on BP Meds", "On BP Meds"))

#prevalentStroke
ggplot(men_data, aes(x = "", fill = as.factor(prevalentStroke))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Men with a History of Stroke") + theme_void() + scale_fill_discrete(name = "People", labels = c("No History of Stroke", "History of Stroke"))
ggplot(women_data, aes(x = "", fill = as.factor(prevalentStroke))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Women with a History of Stroke") + theme_void() + scale_fill_discrete(name = "People", labels = c("No History of Stroke", "History of Stroke"))

#prevalentHyp
ggplot(men_data, aes(x = "", fill = as.factor(prevalentHyp))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Men with a History of Hypertension") + theme_void() + scale_fill_discrete(name = "People", labels = c("No History of Hypertension", "History of Hypertension"))
ggplot(women_data, aes(x = "", fill = as.factor(prevalentHyp))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Women with a History of Hypertension") + theme_void() + scale_fill_discrete(name = "People", labels = c("No History of Hypertension", "History of Hypertension"))

#diabetes
ggplot(men_data, aes(x = "", fill = as.factor(diabetes))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Men with Diabetes") + theme_void() + scale_fill_discrete(name = "People", labels = c("No Diabetes", "Diabetes"))
ggplot(women_data, aes(x = "", fill = as.factor(diabetes))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Women with Diabetes") + theme_void() + scale_fill_discrete(name = "People", labels = c("No Diabetes", "Diabetes"))

#totChol
ggplot(men_data, aes(x = totChol)) + geom_density(fill = "orange", alpha = 0.5) + ggtitle("Distribution of the Total Cholesterol for Men") + xlab("Total Cholesterol")
ggplot(women_data, aes(x = totChol)) + geom_density(fill = "purple", alpha = 0.5) + ggtitle("Distribution of the Total Cholesterol for Women") + xlab("Total Cholesterol")

#sysBP
ggplot(men_data, aes(x = sysBP)) + geom_density(fill = "lightblue", alpha = 0.6) + ggtitle("Distribution of Systolic BP for Men") + xlab("Systolic Blood Pressure")
ggplot(women_data, aes(x = sysBP)) + geom_density(fill = "yellow", alpha = 0.6) + ggtitle("Distribution of Systolic BP for Women") + xlab("Systolic Blood Pressure")

#diaBP
ggplot(men_data, aes(x = diaBP)) + geom_density(fill = "lightgreen", alpha = 0.5) + ggtitle("Distribution of Diastolic BP for Men") + xlab("Diastolic Blood Pressure")
ggplot(women_data, aes(x = diaBP)) + geom_density(fill = "orangered", alpha = 0.5) + ggtitle("Distribution of Diastolic BP for Women") + xlab("Diastolic Blood Pressure")

#BMI
ggplot(men_data, aes (x = BMI)) + geom_histogram(fill = "brown", alpha = 0.9, binwidth = 2) + ggtitle("Distribution of Men BMI")
ggplot(women_data, aes (x = BMI)) + geom_histogram(fill = "gray", alpha = 0.9, binwidth = 2) + ggtitle("Distribution of Women BMI")

#heartRate
ggplot(men_data, aes(x = heartRate)) + geom_density(fill = "purple", alpha = 0.6) + ggtitle("Distribution of Heart Rate for Men") + xlab("Heart Rate")
ggplot(women_data, aes(x = heartRate)) + geom_density(fill = "pink", alpha = 0.6) + ggtitle("Distribution of Heart Rate for Women") + xlab("Heart Rate")

#glucose
ggplot(men_data, aes(x = glucose)) + geom_density(fill = "green", alpha = 0.6) + ggtitle("Distribution of Glucose for Men") + xlab("Glucose")
ggplot(women_data, aes(x = glucose)) + geom_density(fill = "blue", alpha = 0.6) + ggtitle("Distribution of Glucose for Women") + xlab("Glucose")

#tenYearCHD

ggplot(men_data, aes(x = "", fill = as.factor(TenYearCHD))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Men That Developed Coronary Heart Disease") + theme_void() + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))
ggplot(women_data, aes(x = "", fill = as.factor(TenYearCHD))) + geom_bar(position = "stack") + coord_polar("y", start=0) + ggtitle("Distribution of Women That Developed Coronary Heart Disease") + theme_void() + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))


``` 

## AGE FOR MEN AND WOMEN
```{r}

# combined histogram of the ages of men that got chd and those that did not

library(ggplot2)

ggplot(men_data, aes(x = age, fill = as.factor(TenYearCHD))) + geom_histogram(alpha = 0.2) + ggtitle("Distribution of the Ages of Men That Got and Didn't Get CHD") + xlab("Age") + ylab("Count") + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))

ggplot(women_data, aes(x = age, fill = as.factor(TenYearCHD))) + geom_histogram(alpha = 0.2) + ggtitle("Distribution of the Ages of Women Who Got and Didn't Get CHD") + xlab("Age") + ylab("Count") + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))

men_final = read.csv("menfinal.csv")
f = men_final[,-1] # remove first column
convert = as.numeric(f[1,]) #convert first row into a column to put into dataframe
df = data.frame(
  c = 0:(length(convert)-1),
  d = convert
)

ggplot(df, aes(x = c, y = d)) + geom_line() + geom_point() + theme(text=element_text(family="Times New Roman", size=16))

women_final = read.csv("womenfinal.csv")
f1 = women_final[,-1]
convert1 = as.numeric(f1[1,])
wdf = data.frame(c = 0:70, d = convert1)

ggplot(wdf, aes(x = c, y = d)) + geom_line() + geom_point() + theme(text=element_text(family="Times New Roman", size=16))

```

## TOTAL CHOLESTEROL FOR MEN

```{r}
library(ggplot2)
library(geomtextpath)
library(hrbrthemes)

#black and white with a legend and title

ggplot(men_data, aes(x = totChol, linetype = as.factor(TenYearCHD))) + geom_density(alpha = 0.5) + ggtitle("Distribution of the Total Cholesterol for Men Who Got and Didn't Get CHD") + xlab("Total Cholesterol") + scale_linetype_manual(name = "People", labels = c ("No CHD", "CHD"), values = c("solid", "dashed")) + theme_bw() + guides(color = 'none') + theme(legend.position = c(0.9, 0.8))

custom_labels = c("0" = "No CHD", "1" = "CHD")

men_data$custom_label = ifelse(men_data$TenYearCHD == 0, custom_labels[1], custom_labels[2])

#colored, no legend

ggplot(men_data, aes(x = totChol, colour = as.factor(TenYearCHD), label = custom_label)) + geom_density() + geom_textdensity(vjust = -0.1) + theme_bw() + scale_colour_manual(values = c("0" = "blue", "1" = "red"), labels = custom_labels) + guides(color = 'none')
```


## SYSTOLIC BLOOD PRESSURE FOR MEN AND WOMEN

```{r}
library(ggplot2)
library(geomtextpath)
library(hrbrthemes)

#colored, with title

ggplot(men_data, aes(x = sysBP, fill = as.factor(TenYearCHD))) + geom_density(alpha = 0.6) + ggtitle("Distribution of Systolic BP for Men Who Got and Didn't Get CHD") + xlab("Systolic Blood Pressure") + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))

ggplot(women_data, aes(x = sysBP, fill = as.factor(TenYearCHD))) + geom_density(alpha = 0.6) + ggtitle("Distribution of Systolic BP for Women Who Got and Didn't Get CHD") + xlab("Systolic Blood Pressure") + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))

#black and white, no title

ggplot(men_data, aes(x = sysBP, linetype = as.factor(TenYearCHD))) + geom_density(alpha = 0.5) + ggtitle("Distribution of Systolic BP for Men Who Got and Didn't Get CHD") + xlab("Systolic Blood Pressure") + scale_linetype_manual(name = "People", labels = c ("No CHD", "CHD"), values = c("solid", "dashed")) + theme_bw() + guides(color = 'none') + theme(legend.position = c(0.9, 0.8))

ggplot(women_data, aes(x = sysBP, linetype = as.factor(TenYearCHD))) + geom_density(alpha = 0.5) + ggtitle("Distribution of Systolic BP for Women Who Got and Didn't Get CHD") + xlab("Systolic Blood Pressure") + scale_linetype_manual(name = "People", labels = c ("No CHD", "CHD"), values = c("solid", "dashed")) + theme_bw() + guides(color = 'none') + theme(legend.position = c(0.9, 0.8))
```

## GLUCOSE FOR MEN AND WOMEN

```{r}
library(ggplot2)

#colored, with title

ggplot(men_data, aes(x = glucose, fill = as.factor(TenYearCHD))) + geom_density(alpha = 0.6) + ggtitle("Distribution of Glucose for Men Who Got and Didn't Get CHD") + xlab("Glucose") + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))

ggplot(women_data, aes(x = glucose, fill = as.factor(TenYearCHD))) + geom_density(alpha = 0.6) + ggtitle("Distribution of Glucose for Women Who Got and Didn't Get CHD") + xlab("Glucose") + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))

#black and white, no title

ggplot(men_data, aes(x = glucose, linetype = as.factor(TenYearCHD))) + geom_density(alpha = 0.5) + scale_linetype_manual(name = "People", labels = c ("No CHD", "CHD"), values = c("solid", "dashed")) + theme_bw() + guides(color = 'none') + theme(legend.position = c(0.8, 0.8))

ggplot(women_data, aes(x = glucose, linetype = as.factor(TenYearCHD))) + geom_density(alpha = 0.5) + scale_linetype_manual(name = "People", labels = c ("No CHD", "CHD"), values = c("solid", "dashed")) + theme_bw() + guides(color = 'none') + theme(legend.position = c(0.8, 0.8))
```

## CIGS PER DAY FOR WOMEN

```{r}

library(ggplot2)

ggplot(women_data, aes (x = cigsPerDay, fill = as.factor(TenYearCHD))) + geom_histogram(alpha = 0.6, binwidth = 4) + ggtitle("Distribution of The Amount of Cigarettes Smoked by Female Smokers Who Got and Didn't Get CHD") + xlab("Cigarettes Per Day") + ylab("Count") + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD"))


women_cigs = read.csv("womencigs.csv")
f = women_cigs[,-1] # remove first column
convert = as.numeric(f[1,]) #convert first row into a column to put into dataframe
df1 = data.frame(
  c = 0:(length(convert)-1),
  d = convert
)

ggplot(df, aes(x = c, y = d)) + geom_line() + geom_point() + xlab("Cigarettes Per Day") + ylab("Percent That Got a CHD") + ggtitle("Women That Got a CHD As Number of Cigarettes Smoked Per Day Increases")

```

## PREVIOUS HISTORY OF HYPERTENSION FOR WOMEN

```{r}

library(ggplot2)
library(ggpattern)

#colored

ggplot(women_data, aes (x = as.factor(prevalentHyp), fill = as.factor(TenYearCHD))) + geom_bar(alpha = 0.6, width = 0.5) + ggtitle("History of Hypertension in Relation to The Development of a CHD in Women") + scale_fill_discrete(name = "People", labels = c("No CHD", "CHD")) + xlab("History of Hypertension")

#black and white

ggplot(women_data, aes(x=as.factor(prevalentHyp), fill= as.factor(TenYearCHD))) + geom_bar(alpha=.6, width=.5) + scale_fill_grey(start=0, end=0.8) +  theme_bw()

```

## Linear regression model

```{r}
women_model = lm(formula = TenYearCHD ~ age + education + currentSmoker + cigsPerDay + BPMeds + prevalentStroke + prevalentHyp + diabetes + totChol + sysBP + diaBP + BMI + heartRate + glucose, data = women_data)
summary(women_model)

men_model = lm(formula = TenYearCHD ~ age + education + currentSmoker + cigsPerDay + BPMeds + prevalentStroke + prevalentHyp + diabetes + totChol + sysBP + diaBP + BMI + heartRate + glucose, data = men_data)
summary(men_model)
```

## Forward Selection (women)

```{r}
forward1 = lm(formula = TenYearCHD ~ 1, data = women_data)

forward_women = step(forward1, direction = "forward", scope = ~ age + education + currentSmoker + cigsPerDay + BPMeds + prevalentStroke + prevalentHyp + diabetes + totChol + sysBP + diaBP + BMI + heartRate + glucose, data = women_data)

```

## Forward Selection (men)

```{r}
forward2 = lm(formula = TenYearCHD ~ 1, data = men_data)

forward_men = step(forward2, direction = "forward", scope = ~ age + education + currentSmoker + cigsPerDay + BPMeds + prevalentStroke + prevalentHyp + diabetes + totChol + sysBP + diaBP + BMI + heartRate + glucose, data = men_data)

```

## Backward Selection (women)

```{r}
backward1 = lm(formula = TenYearCHD ~ age + education + currentSmoker + cigsPerDay + BPMeds + prevalentStroke + prevalentHyp + diabetes + totChol + sysBP + diaBP + BMI + heartRate + glucose, data = women_data)

backward_women = step(backward1, direction = "backward")

```

## Backward Selection (men)

```{r}
backward2 = lm(formula = TenYearCHD ~ age + education + currentSmoker + cigsPerDay + BPMeds + prevalentStroke + prevalentHyp + diabetes + totChol + sysBP + diaBP + BMI + heartRate + glucose, data = men_data)

backward_men = step(backward2, direction = "backward")

```
## Stepwise Selection (women)

```{r}
stepwise1 = lm(formula = TenYearCHD ~ age + education + currentSmoker + cigsPerDay + BPMeds + prevalentStroke + prevalentHyp + diabetes + totChol + sysBP + diaBP + BMI + heartRate + glucose, data = women_data)

stepwise_women = step(stepwise1, direction = "both")

```

## Stepwise Selection (men)

```{r}
stepwise2 = lm(formula = TenYearCHD ~ age + education + currentSmoker + cigsPerDay + BPMeds + prevalentStroke + prevalentHyp + diabetes + totChol + sysBP + diaBP + BMI + heartRate + glucose, data = men_data)

stepwise_men = step(stepwise2, direction = "both")

```

## women interpretation

```{r}
mod = lm(formula = TenYearCHD ~ age + education + cigsPerDay + BPMeds + prevalentHyp + sysBP + diaBP + glucose, data = women_data)
summary(mod)
```

## men interpretation

```{r}
mod2 = lm(formula = TenYearCHD ~ age + cigsPerDay +  prevalentStroke + totChol + sysBP + glucose, data = men_data)
summary(mod2)
```





